---
title: "Title"
author: "Christian Oppegård Moen"
date: "DD MM YYYY"
output: 
  bookdown::pdf_document2:
    toc_depth: '3'
    number_sections: false
  # pdf_document:
  # #   toc: no
  #   toc_depth: '3'
subtitle: Course
urlcolor: blue
editor_options: 
  chunk_output_type: console
header-includes:
- \usepackage[width=0.8\textwidth]{caption}
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = T, tidy=T, message=F, warning=F,
  strip.white=F,
  prompt=F,
  cache=T,
  root.dir = "./Exe3bsm",
  size="scriptsize",
  fig.width=7, 
  fig.height=5, 
  fig.align = "center"
)
```

```{r load, options}
source("./additionalFiles/probAhelp.R")
source("./additionalFiles/probAdata.R")
figPath = "./Figures/"
```

## 2.



In this problem we want to implement the EM-algorithm. We have found the conditional expectation $Q(\lambda_0,\lambda_1)=Q(\lambda_0, \lambda_1 | \lambda_0^{(t)}, \lambda_1^{(t)}).$ This corresponds to the E-step in the EM algorithm. In the M-step of the algorithm is to determine 

$$
(\lambda_0^{(t+1)}, \lambda_1^{(t+1)})=\text{argmax} \ \ Q(\lambda_0, \lambda_1).
$$

This can be found be finding the partial derivates and $Q(\lambda_0, \lambda_1)$ and set them equal to zero. 

$$
\frac{\partial}{\partial \lambda_0} Q(\lambda_0, \lambda_1)=
\frac{n}{\lambda_0}-\sum_{i=1}^{n} \left ( u_i z_i+(1-u_i) \left ( \frac{1}{\lambda_0^{(t)}}-\frac{z_i}{e^{\lambda_0^{(t)}z_i}-1} \right ) \right)=0
$$

$$
\frac{\partial}{\partial \lambda_1} Q(\lambda_0, \lambda_1)=
\frac{n}{\lambda_1}-\sum_{i=1}^{n} \left ( (1-u_i) z_i+u_i \left ( \frac{1}{\lambda_1^{(t)}}-\frac{z_i}{e^{\lambda_1^{(t)}z_i}-1} \right ) \right)=0
$$

We solve these two equations for $\lambda_0$ and $\lambda_1$ respectively. This gives the M-step

$$
\lambda_0^{(t+1)}=n/\sum_{i=1}^{n} \left ( u_i z_i+(1-u_i) \left ( \frac{1}{\lambda_0^{(t)}}-\frac{z_i}{e^{\lambda_0^{(t)}z_i}-1} \right ) \right)
$$
$$
\lambda_1^{(t+1)}=n/\sum_{i=1}^{n} \left ( (1-u_i) z_i+u_i \left ( \frac{1}{\lambda_1^{(t)}}-\frac{z_i}{e^{\lambda_1^{(t)}z_i}-1} \right ) \right)
$$
Let $\lambda^{(t)}=(\lambda_0^{(t)}, \lambda_1^{(t)}).$
We want to implement the EM-algorithm and we use the convergence criterion

$$
d(x^{(t+1)}, x^{t})= || {\lambda}^{(t+1)} - {\lambda}^{(t)}||_2<\epsilon.
$$

```{r dataC, include=FALSE}
u<-read.csv("./additionalFiles/u.txt")
z<-read.csv("./additionalFiles/z.txt")
```

The function below returns the conditional expectation, that is the E-step of the EM algorithm.

```{r}
cond_expectation <- function(lambda0, lambda1,lambda0t,lambda1t, u, z)
{
  n=length(u)
  exp=n*(log(lambda0)+log(lambda1))-(lambda0*sum(u*z+(1-u)*(1/lambda0t-(z)/(exp(lambda0t*z)-1))))-(lambda1*sum(u*z+(1-u)*(1/lambda1t-(z)/(exp(lambda1t*z)-1))))
return(exp)
}
```

Under is a function that implement M-step.
```{r}
M_step<- function(lambda0t,lambda1t,u,z)
{
  n=length(u)
  lambda0next=n/(sum(u*z+(1-u)*(1/lambda0t-(z)/(exp(lambda0t*z)-1))))
  lambda1next=n/(sum((1-u)*z+u*(1/lambda1t-(z)/(exp(lambda1t*z)-1))))
  
  return(list(lambda0=lambda0next, lambda1=lambda1next))
}

lambda0=2
lambda1=3
list0<-c()
list1<-c()
for (i in 1:10) {
  lambda0=M_step(lambda0,lambda1,u,z)$lambda0
  lambda1=M_step(lambda0,lambda1,u,z)$lambda1
  list0<-c(list0,lambda0)
  list1<-c(list1,lambda1)
}
list0
```

```{r}
EM_algorithm<- function(lambda,u,z, epsilon=10e-5)
{
  lambda0=lambda[1]
  lambda1=lambda[2]
  list0<-c()
  list1<-c()
  for (i in 1:30) {
    lambda0=M_step(lambda0,lambda1,u,z)$lambda0
    lambda1=M_step(lambda0,lambda1,u,z)$lambda1
    list0<-c(list0,lambda0)
    list1<-c(list1,lambda1)
}
  return(list(lambdas0=list0, lambdas1=list1))
}
```


Under the EM algorithm is implemented

```{r}
ME_algorithm<- function(lambda,u,z, epsilon=10e-5)
{
  i=0
  lambda0=lambda[0]
  lambda1=lambda[1]
  lambda=c(lambda0,lambda1)
  lambdas0=c()
  lambdas1=c()
  norm=-Inf
  for(i in 1:50)
  {
    #M-step
    lambda0t=M_step(lambda0,lambda1,u,z)$lambda0
    lambda1t=M_step(lambda0,lambda1,u,z)$lambda1
    lambdas0=c(lambdas0,lambda0t)
    lambdas1=c(lambdas1,lambda1t)
    lambdat=c(lambda0t,lambda1t)
    #convergence
    
    norm=norm(lambdat-lambda, type="2")
    lambda=lambdat
  }
  return(list(lambdas0=lambdas0, lambdas1=lambdas1))
}
```


```{r}
lambdas<-EM_algorithm(c(2.5,5),u,z)
lambdas$lambdas1
```
## 4.

We want to find an analytical formula for $f_{Z_i,U_i}(z_i, u_i| \lambda_0, \lambda_1).$

$$
f_{Z_i,U_i}(z_i, u_i | \lambda_0, \lambda_1)=P(\text{max}(X_i, Y_i)=z , I(X_i \geq Y_i)=u_i | \lambda_0, \lambda_1)
$$
$$
=u_i P(\text{max}(X_i, Y_i)=z_i, X)
$$



