---
title: "Exercise 3"
author: "Erling Fause Steen, Christian Oppegård Moen"
date: "Spring 2022"
output: 
  bookdown::pdf_document2:
    toc_depth: '3'
    number_sections: false
  # pdf_document:
  # #   toc: no
  #   toc_depth: '3'
subtitle: TMA4300
urlcolor: blue
editor_options: 
  chunk_output_type: console
header-includes:
- \usepackage[width=0.8\textwidth]{caption}
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = T, tidy=T, message=F, warning=F,
  strip.white=F,
  prompt=F,
  cache=T,
  root.dir = "./Exe3bsm",
  size="scriptsize",
  fig.width=7, 
  fig.height=5, 
  fig.align = "center"
)
```
\newpage
```{r config, include=F}
if (F){
  setwd("C:/Users/chris/OneDrive/Documents/Fysmat/8 Semester V2022/Ber. Stat. Met/TMA4300/Exe3bsm/")
  options(error=recover)
  options(error=NULL)
}
```

```{r packages, include=F}
library(knitr)
library(kableExtra)
```

# Introduction

# Problem A
In this problem we will analyze a dataset which contain a sequence of length $T = 100$ of a non-Gaussian time-series for which we will compare two different parameter estimators. Consider the $\mathrm{AR}(2)$ specified by the relation
\begin{equation}
  x_t = \beta_1 x_{t-1} + \beta_2 x_{t-2} + e_t,
\end{equation}
where $e_t$ are independent and identically distributed (iid) random variables with zero mean and constant variance. Also, consider the loss functions with respect to $\boldsymbol{\beta} = [\beta_1, \beta_2]^T$ given by
$$
\begin{aligned}
Q_{L S}(\boldsymbol{x})
&=\sum_{t=3}^{T}\left(x_{t}-\beta_{1} x_{t-1}-\beta_{2} x_{t-2}\right)^{2}
\\
Q_{L A}(\boldsymbol{x})
&=\sum_{t=3}^{T}\left|x_{t}-\beta_{1} x_{t-1}-\beta_{2} x_{t-2}\right|.
\end{aligned}
$$
Then, the least sum residuals (LS) and least sum of absolute residuals (LA) are obtained by minimizing $Q_{LS}(\boldsymbol{x})$ and $Q_{L A}(\boldsymbol{x})$ respectively. We denote the minimisers by $\hat{\boldsymbol{\beta}}_{LS}$ and $\hat{\boldsymbol{\beta}}_{LA}$, and define the estimated residuals by $\hat e = x_{t}-\hat{\beta}_{1} x_{t-1}-\hat{\beta}_{2} x_{t-2}$ for $t=3,...,T$ with mean $\bar e$. Then, $\hat \varepsilon = \hat e - \bar e$ is re-centered to have mean zero.

## 1 {#sec:1}
Now we will use the residual resampling bootstrap method to evaluate the relative performance of the two parameter estimators, $\hat{\boldsymbol{\beta}}_{LS}$ and $\hat{\boldsymbol{\beta}}_{LA}$, which are calculated by the given function `ARp.beta.est`. We consider the variance and mean of $\boldsymbol{\beta}^*_{LS}$ and $\boldsymbol{\beta}^*_{LA}$ obtained by minimizing $Q_{LS}(\boldsymbol{x}^*)$ and $Q_{L A}(\boldsymbol{x}^*)$ for bootstrap sample $\boldsymbol{x}^*$, respectively. In Table \ref{tab:A1boot} the variance of $\boldsymbol{\beta}_{LA}$ is slightly lower than that of $\boldsymbol{\beta}_{LS}$, and the absolute value of the bias is slightly lower as well. This suggest that the LS estimator is not optimal.

```{r load, include=F}
source("./additionalFiles/probAhelp.R")
source("./additionalFiles/probAdata.R")
figPath = "./Figures/"
```

```{r disipaly, fig.cap="Given dataset. write more"}
x = data3A$x
plot(x, type="l", xlab = "t", ylab = "x")
```

```{r A1boot, fig.cap="a1tab"}
detach("package:dplyr", unload = T)
rsBoot = function(B, x, p=2){
  T = length(x)
  # Estimate beta
  beta.hat = ARp.beta.est(x,p)
  # calculate observed residuals of AR(p) seq 
  e.LS.observed = ARp.resid(x, beta.hat$LS)
  e.LA.observed = ARp.resid(x, beta.hat$LA)
  # Allocate memory
  beta.LS.star = matrix(nrow = B, ncol = 2)
  beta.LA.star = matrix(nrow = B, ncol = 2)
  # e.LS.star = vector(mode = "double", length = B)
  # x.LS.star = matrix(nrow = B, ncol = 2)
  # x.LA.star = matrix(nrow = B, ncol = 2)
  
  # Bootstrap
  for (b in 1:B){
    # Resample from observed residuals to yield pseudo innovations
    e.LS.star = sample(e.LS.observed, size=T, replace=TRUE)
    e.LA.star = sample(e.LA.observed, size=T, replace=TRUE)
    # Generate pseudo data (Timeseries based on sampled residuals and beta.hat)
    i = sample(T-1, 1)
    x.LS.star = ARp.filter(x[c(i, i+1)], beta.hat$LS, e.LS.star)
    x.LA.star = ARp.filter(x[c(i, i+1)], beta.hat$LA, e.LA.star)
    # compute beta star
    beta.LS.star[b,] = ARp.beta.est(x.LS.star,2)$LS
    beta.LA.star[b,] = ARp.beta.est(x.LA.star,2)$LA
  }
  return(list(beta.hat = beta.hat, 
              beta.LS = beta.LS.star, beta.LA = beta.LA.star,
              e.LS = e.LS.observed, e.LA = e.LA.observed,
              x.LS = x.LS.star, x.LA = x.LA.star))
}

B = 1500
set.seed(420)
boot = rsBoot(B, x)

# Compute variance and bias

beta.var = rbind(apply(boot$beta.LS, 2, var), apply(boot$beta.LA, 2, var))
beta.bias = rbind(apply(boot$beta.LS, 2, mean) - boot$beta.hat$LS,
                  apply(boot$beta.LA, 2, mean) - boot$beta.hat$LA)
beta.vb = cbind(beta.var, beta.bias)
beta.vb
rownames(beta.vb) = c("LS", "LA")
# colnames(beta.vb) = c("var_1", "var_2", "bias_1", "bias_2")

kable(round(beta.vb,6), 
      caption = "Variance and mean for $\\beta_1^*$ and $\\beta_2^*$.", 
      format = "latex") %>% 
  kable_styling() %>%
  add_header_above(c(" ", "Var$_{1}$","Var$_{2}$","bias$_{1}$", 
                     "bias$_{2}$"), escape = FALSE)
```

## 2
Next, we will compute a $95\%$ prediction interval for $x_{101}$ for both the LS and the LA estimator. That is, we bootstrap sample $B$ times
\begin{equation}
  x_{101} = \beta_1^* x_{100} + \beta_2^* x_{99} + \hat{e}_{i}
\end{equation}
where $\beta_1^*$ and $\beta_2^*$ are sampled from all 1500 bootstrap samples found in section [A1](#sec:1), and $\hat{e}_{i}$ is sampled from the estimated residuals $\hat{e}_{t}$ for $t = 2,...,100$. The prediction intervals are found in Table \ref{tab:A1pi}.
```{r A1pi, fig.cap="a1pi"}
set.seed(420)
# Bootstrap x_101 using beta.star and e.observed
x_101.LS = vector(mode = "double", length = B)
x_101.LA = vector(mode = "double", length = B)
for (b in 1:B){
  i.e = sample(98, 1)
  i.beta = sample(1500,1)
  x_101.LS[b] = boot$beta.LS[i.beta,] %*% x[c(100,99)] + boot$e.LS[i.e]
  x_101.LA[b] = boot$beta.LS[i.beta,] %*% x[c(100,99)] + boot$e.LA[i.e]
}

pi.boot = rbind(LS = quantile(x_101.LS, probs=c(0.025,0.975)),
                LA = quantile(x_101.LA, probs=c(0.025,0.975)))
kable(round(pi.boot,2), caption = "Prediction interval of $x_{101}$.")
```

# Problem B
We will investigate the concentration of bilirubin (mg/dL) in blood samples taken from three young men.

$$
\begin{array}{c|ccccccccccc}
\hline \text { Individual } & {\text { Concentration }(\mathrm{mg} / \mathrm{dL})} \\
\hline 1 & 0.14 & 0.20 & 0.23 & 0.27 & 0.27 & 0.34 & 0.41 & 0.41 & 0.55 & 0.61 & 0.66 \\
2 & 0.20 & 0.27 & 0.32 & 0.34 & 0.34 & 0.38 & 0.41 & 0.41 & 0.48 & 0.55 & \\
3 & 0.32 & 0.41 & 0.41 & 0.55 & 0.55 & 0.62 & 0.71 & 0.91 & & & \\
\hline
\end{array}
$$
```{r display, options}
bilirubin <- read.table("./additionalFiles/bilirubin.txt",header=T)
bilirubin
boxplot(bilirubin)
```