---
title: "Title"
author: "Christian Oppegård Moen"
date: "DD MM YYYY"
output: 
  bookdown::pdf_document2:
    toc_depth: '3'
    number_sections: false
  # pdf_document:
  # #   toc: no
  #   toc_depth: '3'
subtitle: Course
urlcolor: blue
editor_options: 
  chunk_output_type: console
header-includes:
- \usepackage[width=0.8\textwidth]{caption}
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = T, tidy=T, message=F, warning=F,
  strip.white=F,
  prompt=F,
  cache=T,
  root.dir = "./Exe3bsm",
  size="scriptsize",
  fig.width=7, 
  fig.height=5, 
  fig.align = "center"
)
```

```{r load, options}
source("./additionalFiles/probAhelp.R")
source("./additionalFiles/probAdata.R")
source("./additionalFiles/u.txt")
source("./additionalFiles/z.txt")
figPath = "./Figures/"
```

# Problem C: The EM-algorithm and bootstrapping

Let $x_1,...x_n$ and $y_1,...,y_n$ be independet random variables, where
$$
x_i \sim \text{Exp}(\lambda_0) \ \ \text{and} \ \ y_i \sim \text{Exp}(\lambda_1)
$$

We observe

$$
z_i =\text{max}(x_i, y_i) \ \ \text{for} \ \ i=1,...,n
$$

and

$$
u_i=I(x_i \geq y_i) \ \ \text{for} \ \ i=1,...,n.
$$

## 1.

The joint distribution of $(x_i, y_i),i=1,..n$ is given by

$$
f(x, y | \lambda_0, \lambda_1)=\prod_{i=1}^{n} f_{x}(x_i | \lambda_0) \cdot f_{y}(y_i|\lambda_1)
$$
$$
=\prod_{i=1}^{n}\lambda_0 e^{-\lambda_0 x_i} \cdot \lambda_1 e^{-\lambda_1 y_i}.
$$
This means that the log likelihood is given by
$$
\ln f(x,y|\lambda_0, \lambda_1)=\sum_{i=1}^{n} \ln \lambda_0+ \ln \lambda_1-\lambda_0 x_i - \lambda_1 y_i.
$$
We want to find
$$
E\left[ \ln f(x,y|\lambda_0, \lambda_1)| z, u, \lambda_0^{(t)}, \lambda_1^{(t)} \right].
$$

which is given by

$$
Q(\lambda_0, \lambda_1 | \lambda_0^{(t)}, \lambda_1^{(t)})=E \left[\sum_{i=1}^{n}( \ln \lambda_0+ \ln \lambda_1-\lambda_0 x_i - \lambda_1 y_i) | z,u,\lambda_0^{(t)}, \lambda_1^{(t)} \right]
$$
$$
=n(\ln \lambda_0+ \ln \lambda_1)-\lambda_0 \sum_{i=1}^{n} E
$$

## 2.

In this problem we want to implement the EM-algorithm. We have found the conditional expectation $Q(\lambda_0,\lambda_1)=Q(\lambda_0, \lambda_1 | \lambda_0^{(t)}, \lambda_1^{(t)}).$ This corresponds to the E-step in the EM algorithm. In the M-step of the algorithm is to determine 

$$
(\lambda_0^{(t+1)}, \lambda_1^{(t+1)})=\text{argmax} \ \ Q(\lambda_0, \lambda_1).
$$

This can be found be finding the partial derivates and $Q(\lambda_0, \lambda_1)$ and set them equal to zero. 

$$
\frac{\partial}{\partial \lambda_0} Q(\lambda_0, \lambda_1)=
\frac{n}{\lambda_0}-\sum_{i=1}^{n} \left ( u_i z_i+(1-u_i) \left ( \frac{1}{\lambda_0^{(t)}}-\frac{z_i}{e^{\lambda_0^{(t)}z_i}-1} \right ) \right)=0
$$

$$
\frac{\partial}{\partial \lambda_1} Q(\lambda_0, \lambda_1)=
\frac{n}{\lambda_1}-\sum_{i=1}^{n} \left ( u_i z_i+(1-u_i) \left ( \frac{1}{\lambda_1^{(t)}}-\frac{z_i}{e^{\lambda_1^{(t)}z_i}-1} \right ) \right)=0
$$

We solve these two equations for $\lambda_0$ and $\lambda_1$ respectively. This gives the M-step

$$
\lambda_0^{(t+1)}=n/\sum_{i=1}^{n} \left ( u_i z_i+(1-u_i) \left ( \frac{1}{\lambda_0^{(t)}}-\frac{z_i}{e^{\lambda_0^{(t)}z_i}-1} \right ) \right)
$$
$$
\lambda_1^{(t+1)}=n/\sum_{i=1}^{n} \left ( u_i z_i+(1-u_i) \left ( \frac{1}{\lambda_1^{(t)}}-\frac{z_i}{e^{\lambda_1^{(t)}z_i}-1} \right ) \right)
$$
Let $\lambda^{(t)}=(\lambda_0^{(t)}, \lambda_1^{(t)}).$
In the code below ,the EM-agorithm is implemented with the convergence criterion

$$
d(x^{(t+1)}, x^{t})= || {\lambda}^{(t+1)} - {\lambda}^{(t)}||_2<\epsilon
$$



