---
title: "Exercise 3"
author: "Erling Fause Steen, Christian Oppegård Moen"
date: "Spring 2022"
output: 
  bookdown::pdf_document2:
    toc_depth: '3'
    number_sections: false
  # pdf_document:
  # #   toc: no
  #   toc_depth: '3'
subtitle: TMA4300
urlcolor: blue
editor_options: 
  chunk_output_type: console
header-includes:
- \usepackage[width=0.8\textwidth]{caption}
- \usepackage{algorithm}
- \usepackage{algpseudocode}
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = T, tidy=T, message=F, warning=F,
  strip.white=F,
  prompt=F,
  cache=T,
  root.dir = "./Exe3bsm",
  size="scriptsize",
  fig.width=7, 
  fig.height=5, 
  fig.align = "center"
)
```
\newpage
```{r config, include=F}
if (F){
  setwd("C:/Users/chris/OneDrive/Documents/Fysmat/8 Semester V2022/Ber. Stat. Met/TMA4300/Exe3bsm/")
  options(error=recover)
  options(error=NULL)
}
```

```{r packages, include=F}
library(knitr)
library(kableExtra)
```

# Introduction

# Problem A
In this problem we will analyze a dataset which contain a sequence of length $T = 100$ of a non-Gaussian time-series displayed in Figure \ref{fig:Adisiplay}, for which we will compare two different parameter estimators. Consider the autoregressive model of order 2 ($\mathrm{AR}(2)$) specified by the relation
\begin{equation*}
  x_t = \beta_1 x_{t-1} + \beta_2 x_{t-2} + e_t,
\end{equation*}
where $e_t$ are independent and identically distributed (iid) variables with zero mean and constant variance. Also, consider the loss functions with respect to $\boldsymbol{\beta} = [\beta_1, \beta_2]^T$ given by
$$
\begin{aligned}
Q_{L S}(\boldsymbol{x})
&=\sum_{t=3}^{T}\left(x_{t}-\beta_{1} x_{t-1}-\beta_{2} x_{t-2}\right)^{2}
\\
Q_{L A}(\boldsymbol{x})
&=\sum_{t=3}^{T}\left|x_{t}-\beta_{1} x_{t-1}-\beta_{2} x_{t-2}\right|.
\end{aligned}
$$
Then, the least sum residuals (LS) and least sum of absolute residuals (LA) are obtained by minimizing $Q_{LS}(\boldsymbol{x})$ and $Q_{L A}(\boldsymbol{x})$ respectively. We denote the minimisers (our two different parameter estimators) by $\hat{\boldsymbol{\beta}}_{LS}$ and $\hat{\boldsymbol{\beta}}_{LA}$, and define the estimated residuals by $\hat e_t = x_{t}-\hat{\beta}_{1} x_{t-1}-\hat{\beta}_{2} x_{t-2}$ for $t=3,...,T$ with mean $\bar e$. 
<!-- Then, $\hat \varepsilon = \hat e - \bar e$ is re-centered to have mean zero. -->

```{r load, include=F}
source("./additionalFiles/probAhelp.R")
source("./additionalFiles/probAdata.R")
figPath = "./Figures/"
```

```{r Adisiplay, fig.cap="The non-Gaussian time-series.", fig.width=6, fig.height=4, out.width="70%"}
x = data3A$x
plot(x, type="l", xlab = "t", ylab = "x")
```

## 1 {#sec:1}
Now we will use the residual resampling bootstrap method to evaluate the relative performance of the two parameter estimators, $\hat{\boldsymbol{\beta}}_{LS}$ and $\hat{\boldsymbol{\beta}}_{LA}$, which are calculated by the given function `ARp.beta.est` in `R`. We consider the variance and bias of the estimators $\hat{\boldsymbol{\beta}}^*_{LS}$ and $\hat{\boldsymbol{\beta}}^*_{LA}$ obtained by minimizing $Q_{LS}(\boldsymbol{x}^*)$ and $Q_{L A}(\boldsymbol{x}^*)$ respectively, for pseudo data $\boldsymbol{x}^*$. The full method is shown in Algorithm \ref{alg:A1boot}. In Table \ref{tab:A1boot} the variance of $\hat{\boldsymbol{\beta}}^*_{LA}$ is lower than that of $\hat{\boldsymbol{\beta}}^*_{LS}$, and the absolute value of the bias is lower as well. This suggest that the LS estimator is not optimal for this non-Gaussian time-series. 

\begin{algorithm}
  \caption{}
  \begin{algorithmic}[1]
    \For{LS and LA estimator of $\boldsymbol{\beta}$}
      \For{$t=3,...,n$}
        \State Define estimated residuals $\hat e_t$.
      \EndFor
      \For{$b = 1,...,B$}
        \State Resample $n+1$ pseudo residuals $\hat{ \boldsymbol{e}}_b^*$ from the estimated residuals $\hat{ \boldsymbol{e}}_b$ with replacement.
        \State Generate pseudo data $\boldsymbol{x}^*_b$.
        \State Estimate $\hat{\boldsymbol{\beta}}^*_b$.
      \EndFor
    \EndFor
  \end{algorithmic}
  \label{alg:A1boot}
\end{algorithm}


```{r A1boot, fig.cap="a1tab"}
detach("package:dplyr", unload = T)
rsBoot = function(B, x, p=2){
  T = length(x)
  # Estimate beta
  beta.hat = ARp.beta.est(x,p)
  # calculate observed residuals of AR(p) seq 
  e.LS.observed = ARp.resid(x, beta.hat$LS)
  e.LA.observed = ARp.resid(x, beta.hat$LA)
  # Allocate memory
  beta.LS.star = matrix(nrow = B, ncol = 2)
  beta.LA.star = matrix(nrow = B, ncol = 2)
  
  # Bootstrap
  for (b in 1:B){
    # Resample from observed residuals to yield pseudo innovations
    e.LS.star = sample(e.LS.observed, size=T, replace=TRUE)
    e.LA.star = sample(e.LA.observed, size=T, replace=TRUE)
    # Generate pseudo data (Timeseries based on sampled residuals and beta.hat)
    i = sample(T-1, 1)
    x.LS.star = ARp.filter(x[c(i, i+1)], beta.hat$LS, e.LS.star)
    x.LA.star = ARp.filter(x[c(i, i+1)], beta.hat$LA, e.LA.star)
    # compute beta star
    beta.LS.star[b,] = ARp.beta.est(x.LS.star,2)$LS
    beta.LA.star[b,] = ARp.beta.est(x.LA.star,2)$LA
  }
  return(list(beta.hat = beta.hat, 
              beta.LS = beta.LS.star, beta.LA = beta.LA.star,
              e.LS = e.LS.observed, e.LA = e.LA.observed,
              x.LS = x.LS.star, x.LA = x.LA.star))
}

B = 1500
set.seed(420)
boot = rsBoot(B, x)

# Compute variance and bias
beta.var = rbind(apply(boot$beta.LS, 2, var), apply(boot$beta.LA, 2, var))
beta.bias = rbind(apply(boot$beta.LS, 2, mean) - boot$beta.hat$LS,
                  apply(boot$beta.LA, 2, mean) - boot$beta.hat$LA)
beta.vb = cbind(beta.var, beta.bias)
rownames(beta.vb) = c("LS", "LA")

kable(round(beta.vb,6), 
      caption = "Variance and bias for $\\hat{\\beta_1^*}$ and $\\hat{\\beta_2^*}$.", 
      format = "latex") %>% 
  kable_styling() %>%
  add_header_above(c(" ", "Var$_{1}$","Var$_{2}$","bias$_{1}$", 
                     "bias$_{2}$"), escape = FALSE)
```

## 2
Next, we will compute a $95\%$ prediction interval for $x_{101}$ for both the LS and the LA estimator. That is, we bootstrap sample $B$ times
\begin{equation*}
  x_{101} = \beta_1^* x_{100} + \beta_2^* x_{99} + \hat{e}_{i}
\end{equation*}
where $\beta_1^*$ and $\beta_2^*$ are sampled from all `r B` bootstrap samples $\hat{\boldsymbol{\beta}}^*$ found in section [A1](#sec:1), and $\hat{e}_{i}$ is sampled from the estimated residuals $\hat{e}_{t}$ for $t = 2,...,100$. The prediction intervals are found in Table \ref{tab:A1pi}.
```{r A1pi, fig.cap="a1pi"}
set.seed(420)
# Bootstrap x_101 using beta.star and e.observed
x_101.LS = vector(mode = "double", length = B)
x_101.LA = vector(mode = "double", length = B)
for (b in 1:B){
  # i.e = sample(98, 1)
  i.beta = sample(1500,1)
  # x_101.LS[b] = boot$beta.LS[i.beta,] %*% x[c(100,99)] + boot$e.LS[i.e]
  # x_101.LA[b] = boot$beta.LS[i.beta,] %*% x[c(100,99)] + boot$e.LA[i.e]
  x_101.LS[b] = boot$beta.LS[i.beta,] %*% x[c(100,99)] + sample(boot$e.LS,1)
  x_101.LA[b] = boot$beta.LS[i.beta,] %*% x[c(100,99)] + sample(boot$e.LA,1)
}

pi.boot = rbind(LS = quantile(x_101.LS, probs=c(0.025,0.975)),
                LA = quantile(x_101.LA, probs=c(0.025,0.975)))
kable(round(pi.boot,2), caption = "Prediction interval of $x_{101}$.")
```

# Problem B
We will investigate the concentration of bilirubin (mg/dL), which is a breakdown of haemoglobin, in blood samples taken from three young men shown in Table \ref{tab:Bdata}.
\begin{table}
\begin{tabular}{c|ccccccccccc}
\hline \text { Individual } & \multicolumn{11}{c}{\text { Concentration }($\mathrm{mg} / \mathrm{dL})$} \\
\hline 1 & 0.14 & 0.20 & 0.23 & 0.27 & 0.27 & 0.34 & 0.41 & 0.41 & 0.55 & 0.61 & 0.66 \\
2 & 0.20 & 0.27 & 0.32 & 0.34 & 0.34 & 0.38 & 0.41 & 0.41 & 0.48 & 0.55 & \\
3 & 0.32 & 0.41 & 0.41 & 0.55 & 0.55 & 0.62 & 0.71 & 0.91 & & & \\
\hline
\end{tabular}
\caption{Concentration of bilirubin in three different individuals.}
\label{tab:Bdata}
\end{table}

## 1 {#sec:B1}
The logarithms of the concentration for each individual are displayed in the boxplot in Figure \ref{fig:B1display}. We see that the mean logarithm of bilirubin is smallest for individual $1$ and highest for individual $2$. The variation is smallest for individual $2$ while individual $1$ have slightly larger standard deviation than individual $3$.
```{r B1display, fig.cap="The logarithm of the concentration of bilirubin (\\texttt{meas}) for each individual.", fig.width=6, fig.height=4, out.width="70%"}
ylim = c(-2.1, 0)
bilirubin <- read.table("./additionalFiles/bilirubin.txt",header=T)
boxplot(log(meas)~pers,data=as.data.frame(bilirubin), xlab = "Person", ylab = "log(meas)", ylim=ylim)
```

```{r B1linFit, options}
fitB1 = lm(log(meas)~pers, data=bilirubin)
fitB1.Fstat = summary(fitB1)$fstatistic
Fval = fitB1.Fstat["value"]
Pval = pf(Fval, fitB1.Fstat[2], fitB1.Fstat[3], lower.tail = F)
```
Let $\mathrm{log}Y_{ij}=\mathrm{log}(\texttt{meas}_{ij})$ for individual $i=1,2,3$ and observation $j=1,...,n_i$, where $n_1=11$, $n_2=10$ and $n_3=8$. In the chunk above we fit the regression model
\begin{equation}
  \mathrm{log}Y_{ij}=\beta_i + \varepsilon_{ij},
\end{equation}
where $\varepsilon_{ij} \stackrel{iid}{\sim} \mathcal{N}(0,\sigma^2)$. Then, the F-statistic on $3-1$ and $29-3$ degrees of freedom of the hypothesis that $\beta_1=\beta_2=\beta_3$ is `r round(Fval,2)` with p-value `r round(Pval, 5)`. With the F-statistic we are investigating whether the bilirubin of each individual are significantly equally distributed. Judging from the p-value we reject the hypothesis on a $\alpha=0.05$-level. That is, on a $\alpha = 0.05$-level, the individuals do not share the same concentration of bilirubin.

## 2
To further investigate the individuals bilirubin we will consider a permutation test where the idea is that shuffling the labels will not change the joint distribution of the data. That is, we assume that the distributions are equal so that shuffling the order of the data to generate bootstrap samples should be valid. This is done by first randomly assigning bilirubin values, `meas`,  to the three individuals. then we generate $B=999$ bootstrap samples. Implementation of the permutation and computation of the F-statistic is seen in the chunk below. 

```{r B2perm, options}
permTest = function(data=bilirubin){
  bilirubin$perm = bilirubin$meas[sample(29,29)]
  return(summary(lm(log(perm)~pers, data = bilirubin))$fstatistic["value"])
}
```

## 3 {#sec:B3}
We generate $B=999$ samples of the F-statistic, `Fvals`,  using the function `permTest`. Then we compute the p-value defined by `Pval`$= ($`Fvals`$\geq$`Fval`$)/B$, where `Fval` is the F-statistic found in section [B1](#sec:B1). 
```{r B3run, options}
set.seed(420)
B=999
Fvals = vector(mode = "double", length = 999)
for (b in 1:B){
  Fvals[b] = permTest()
}
Pval.perm = sum(Fvals>=Fval)/B
```
The p-value of the $999$ F-statistics for this run was `r round(Pval.perm,5)` which also reject the hypothesis on a $\alpha=0.05$-level. Comparing it to the p-value from section [B1](#sec:B1) we see that they are almost equal, with the value from the permutation test being slightly lower.