

---
title: "Exercise 2:\\ Problem 2 c) Solution"
author: "Christian Oppegård Moen and Erling Fause Steen"
date: "3/22/2022"
output:
  beamer_presentation:
    slide_level: 2
    theme: "Singapore"
    colortheme: "default"
    font: "serif"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(INLA)
library(ggplot2)
```

```{r}
load("~/Beregningskrevende-statistiske-modeller/TMA4300/Exe2bsm/rain.rda")
```

## Problem 2 c)

In this problem, we are asked to consider the following model in INLA:

```{r}
alpha=2
beta=0.05
hyper=list(theta=list(prior="loggamma", param=c(alpha,beta)))
control.inla=list(strategy="simplified.laplace", int.strategy="ccd")
```

```{r}
mod1<-inla(n.rain ~ -1+f(day, model="rw1", constr=FALSE),
data=rain, Ntrials=n.years, control.compute=list(config = TRUE),
family="binomial", verbose=TRUE, control.inla=control.inla)
```


\small
```{r, echo = T, eval = T}
mod <- inla(n.rain ~ f(day, model="rw1", constr=TRUE),
data=rain, Ntrials=n.years, control.compute=list(config = TRUE),
family="binomial", verbose=TRUE, control.inla=control.inla)
```
\normalsize
## Want to find out:
* How is it different from the model in 2a mathematically?

*  Are the predictions from this model significantly different from the model in 2a?

*  Why?


## Comparison with the model in 2a

* Model in 2a:
$$
y_t | \tau_t \sim \text{Bin}(n_t, \pi(\tau_t)), \ \ \pi(\tau_t)=\frac{\exp(\tau_t)}{1+\exp(\tau_t)}=\frac{1}{1+\exp(-\tau_t)}.
$$

* The new model includes an intercept term:
$$
y_t | \tau_t \sim \text{Bin}(n_t, \pi(\eta_t)), \ \ \pi(\eta_t)=\frac{\exp(\eta_t)}{1+\exp(\eta_t)}=\frac{1}{1+\exp(-\eta_t)}.
$$
where $\eta_t=\alpha+\tau_t$ and $\alpha$ is the intercept term. 

* Default prior on $\alpha$. 

* $\text{Loggamma}(\alpha,\beta)$ prior on $\theta$.

* Constr=TRUE $\implies$ Sum-to-zero constraint:

$$
\sum_{t=1}^{T} \tau_t=0.
$$

## Comparison of predictions of $\tau$

```{r plot-2c-tau, fig.cap="$\\tau$-values of model 1 and 8"}

tau_1=mod1$summary.random$day$mean
tau=mod$summary.random$day$mean
label1=c("Model 1", "Model 8")
ggplot()+
  geom_line(data=as.data.frame(tau_1), mapping=aes(x=1:366, y=tau_1, color="Model without intercept"))+geom_line(data=as.data.frame(tau), mapping=aes(x=1:366, y=tau, color="Model with intercept"))+xlab("Day")+ylab("Predicted values of"~tau)+
  labs(color="Models")
```


## Comparison of predictions of $\pi$

```{r plot-2c, fig.cap="Fitted values of the two models"}
pi_1<-mod1$summary.fitted.values$mean
pi<-mod$summary.fitted.values$mean
ggplot()+
  geom_line(data=as.data.frame(pi_1), mapping=aes(x=1:366, y=pi_1, color="Model without intercept"))+geom_line(data=as.data.frame(pi), mapping=aes(x=1:366, y=pi, color="Model with intercept"))+xlab("Day")+ylab("Predicted values of "~pi)+
  labs(color="Models")
```

## Comparison of predictions of $\pi_{201}$ and $\pi_{366}$

* Predictions from the new model:

\tiny
```{r, echo=F, eval=T}
mod$summary.fitted.values[201,]
mod$summary.fitted.values[366,]
```
\normalsize
* Predictions from model in 2a:

\tiny
```{r, echo=F, eval=T}
mod1$summary.fitted.values[201,]
mod1$summary.fitted.values[366,]
```

## Explanantion

* Different constraints $\implies$ different results for $\tau$
* Same shape 
* Adding the intercept gives almost identical predictions of $\pi$


Mathematically:

* Same posterior distribution 



## Conclusions

>- The model has an intercept term and sum-to-zero constraint

>- No significant differences in predictions between the two models

>- Adding the intercept term to the model with sum-to-zero constraint makes the model as flexible as the model in 2a



## 

\center
\Large
Questions?


