---
title: "Project 2"
author: "Erling Fause Steen og Christian Oppegård Moen"
date: "08 03 2022"
output: 
  bookdown::pdf_document2:
    toc_depth: '3'
    number_sections: false
  # pdf_document:
  # #   toc: no
  #   toc_depth: '3'
subtitle: Computer Intensive Statistical Methods
urlcolor: blue
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,
                      strip.white=TRUE,
                      prompt=FALSE,
                      cache=TRUE,
                      root.dir = "./Exe2bsm",
                      size="scriptsize",
                      fig.width=7, 
                      fig.height=5, 
                      fig.align = "center")
```


```{r imports, include=FALSE}
library(ggplot2)
library(MASS)
library(Matrix)
if (F){
  options(error = recover)
  options(error = NULL)
}
# if (getwd() == "C:/Users/chris/OneDrive/Documents/Fysmat/8 Semester V2022/Ber. Stat. Met/TMA4300"){
#   setwd("./Exe2bsm")
# }
```

# Introduction
The Tokyo rainfall dataset contain the amount of rainfall for each of the $366$ days (including February 29.) for several years. We will consider a portion of this dataset, specifically from $1951-1989$, such that for each day $t\neq 60$ we have $n_t=39$ observations and for $t=60$, February 39, we have $n_t=10$ observation. For each day we have the response $y_t = 0,1,2,...,n_t$ being the amount of times the rainfall exceeded $1$mm over the given period, given by
\begin{equation}(\#eq:1response)
  y_t | \tau_t \sim \text{Bin}(n_t,\pi (\tau _t)), \ \ \pi(\tau_t)=\frac{\exp(\tau_t)}{1+\exp(\tau_t)}=\frac{1}{1+\exp(-\tau_t)}.
\end{equation}
Here, $\pi(\tau_t)$ is the probability of rainfall exceeding $1$mm and $\tau_t$ is the logit probability of exceedence. For this project we assume conditional independence among the $y_t|\tau_t \ \forall t=1,2,...,366$.

We will apply a Bayesian hierarchical model to the dataset, using a random walk of order 1 (RW1) to model the trend. For the model we will implement a Markov chain Monte Carlo (MCMC) sampler for the posterior using Metropolis-Hastings (MH) and Gibbs steps for specific parameters. Then we will investigate the accuracy and computational speed of the implementation compared to the built in method `INLA` in R. 

# Problem 1
## a) Display
In Figure \@ref(fig:displayData) we see the number of times the rainfall has exceeded $1$mm. There seem to be fewer days in the start of the year and in the end of the year with an amount of rainfall exceeding 1 mm. This is in January and December. The number of days steadily increases until the begining of the summer which seems to be the period with the most days with an amount of rainfall over 1 mm. Then, the amount of days decreases during july and august before increasing during the autumn. There also seem to be fluctuations on a daily basis from the just mentiod trend in the data. The red dot is the observation for February 29.

```{r displayData, fig.cap="The Tokyo Rainfall dataset"}
load("./rain.rda")
##Plotting the data
ggplot(data=rain, mapping=aes(x=day, y=n.rain)) + 
  geom_line() + 
  xlab("Day") + 
  ylab("Number of days with more than 1mm rain") + 
  geom_point(aes(x=day[60], y=n.rain[60]), colour="red")
```


## b) Likelihood
The likelihood of Equation \@ref(eq:1response) is given by
$$
\begin{array}{rl}
  L(\pi (\boldsymbol{\tau})) &= \prod_{i=1}^{T} \left(
  \begin{array}{c}
    n_t \\ y_t
  \end{array} \right)
  \pi(\tau_t)^{y_t}(1-\pi(\tau_t)^{n_t-y_t} 
  \\
  &\propto \prod_{i=1}^{T} \pi(\tau_t)^{y_t}(1-\pi(\tau_t)^{n_t-y_t} 
  \\
  &= \prod_{t=1}^{T} \left(\frac{\exp(\tau_t)}{1+\exp(\tau_t)}\right)^{y_t}
  \left( 1- \frac{\exp(\tau_t)}{1+\exp(\tau_t)}\right)^{n_t - y_t},
\end{array}
$$
where $\boldsymbol{\tau} = (\tau_1,..., \tau_T)$, $y_t = 1,2,...,39$ and $n_t = 39$ for $t \neq 60$, and $y_t = 1,2,...,10$ and $n_t = 10$ for $t \neq 60$.

## c) Posterior
As briefly mentioned in the introduction we need the posterior $P(\sigma^2 | \tau, y)$ for the Gibbs step in our implementation, given by
$$
\begin{aligned}
  P(\sigma^2 | \boldsymbol{\tau}, \boldsymbol{y}) &= \frac{P(\sigma^2_u, \boldsymbol{\tau}, \boldsymbol{y})}{P(\boldsymbol{\tau}, \boldsymbol{y})}
  \\
  &\propto  P(\boldsymbol{y} | \sigma^2_u,\boldsymbol{\tau}) P(\sigma^2_u,\boldsymbol{\tau})
  \\
  &= P(\boldsymbol{y} | \sigma^2_u,\boldsymbol{\tau}) P(\tau|\sigma^2_u)P(\sigma^2_u),
\end{aligned}
$$
where $\boldsymbol{y} = (y_1,...,y_T)^T$, $\boldsymbol{\tau} = (\tau_1,..., \tau_T)$ and $P(\boldsymbol{y} | \sigma^2_u,\boldsymbol{\tau}) = L(\pi( \boldsymbol{\tau}))$. Based on model assumptions mentioned in the introduction, we have $\tau_t \sim \tau_{t-1}$ for $u_t \stackrel{iid}{\sim} \mathcal{N}(0, \sigma_u^2)$ so that
$$
p(\boldsymbol{\tau} | \sigma_u^2) = \prod_{t=2}^{T} \frac{1}{\sigma_u} \exp\left\{ -\frac{1}{2\sigma^2_u}(\tau_t - \tau_{t-1})^2 \right\}.
$$
We place an inverse gamma prior (IG) on $\sigma_u^2$ given by
$$
p(\sigma_u^2) = \frac{\beta^\alpha}{\Gamma(\alpha)} \left( \frac{1}{\sigma_u^2} \right)^{\alpha+1} exp\left\{ - \frac{\beta}{\sigma_u^2} \right\}.
$$
Then, the posterior is

$$
\begin{aligned}
  P(\sigma^2 | \boldsymbol{\tau}, \boldsymbol{y}) &= \underbrace{\prod_{t=1}^{T} \left(\frac{\exp(\tau_t)}{1+\exp(\tau_t)}\right)^{y_t}
  \left( 1- \frac{\exp(\tau_t)}{1+\exp(\tau_t)}\right)^{n_t - y_t}}_{\text{Constant w.r.t. }\sigma^2}\cdot \\
  &\quad \ \prod_{t=1}^{T} \frac{1}{\sigma_u} \exp\left\{ -\frac{1}{2\sigma^2_u}(\tau_t - \tau_{t-1})^2 \right\} \cdot \frac{\beta^\alpha}{\Gamma(\alpha)} \left( \frac{1}{\sigma_u^2} \right)^{\alpha+1} exp\left\{ - \frac{\beta}{\sigma_u^2} \right\}
  \\
  &\propto \prod_{t=1}^{T} \frac{1}{\sigma_u} \exp\left\{ -\frac{1}{2\sigma^2_u}(\tau_t - \tau_{t-1})^2 \right\} \cdot \frac{\beta^\alpha}{\Gamma(\alpha)} \left( \frac{1}{\sigma_u^2} \right)^{\alpha+1} exp\left\{ - \frac{\beta}{\sigma_u^2} \right\}
  \\
  &= \frac{1}{\sigma^{T-1}_u} exp\left\{ -\frac{1}{2\sigma^2_u} \boldsymbol{\tau Q \tau} \right\} \cdot \frac{\beta^\alpha}{\Gamma(\alpha)} \left( \frac{1}{\sigma_u^2} \right)^{\alpha+1} exp\left\{ - \frac{\beta}{\sigma_u^2} \right\}
  \\
  &\propto \left( \frac{1}{\sigma_u^2} \right)^{\alpha+\frac{T-1}{2} + 1} 
  exp\left\{ -\frac{1}{\sigma^2_u} \left( \frac12 \boldsymbol{\tau Q \tau} + \beta \right)  \right\}
\end{aligned}
$$
for a tridiagonal matrix $\boldsymbol{Q}$ with diagonal elements equal to $2$ except first and last element which are $1$, and the offdiagonal elements equal to $-1$. We recognize the posterior as the core of an inverse gamma $\mathrm{IG}(\alpha^*, \beta^*)$ with shape $\alpha^* = \alpha + \frac12(T-1)$ and scale $\beta^* = \beta + \frac12 \boldsymbol{\tau Q \tau}$.

## d) Acceptance probability
Let $\mathcal{I} \subseteq \{1,2,...,366\}$ be a set of time indices, and let $- \mathcal{I} = \{1,2,...,366\} \setminus \mathcal{I}$. Furthermore, let $\boldsymbol{\tau}'$ denote the proposed values for $\boldsymbol{\tau}$. The MH step needs an acceptance probability denoted $\alpha$ for the proposed values $\boldsymbol{\tau}'_\mathcal{I}$. By using iterative conditioning we can write the acceptance probability as
$$
 \alpha(\boldsymbol{\tau}_ \mathcal{I}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2, \boldsymbol{y}) 
 = \text{min} \left( 1, \frac{P( \boldsymbol{\tau}'_ \mathcal{I}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2, \boldsymbol{y})}
 {P( \boldsymbol{\tau}_ \mathcal{I}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2, \boldsymbol{y})} 
 \frac{Q( \boldsymbol{\tau}_\mathcal{I}|\boldsymbol{\tau}_{-\mathcal{I}}, \sigma_u^2, \boldsymbol{y})}
 {Q(\boldsymbol{\tau}'_ \mathcal{I}|\boldsymbol{\tau}_{-\mathcal{I}}, \sigma_u^2, \boldsymbol{y})}
 \right),
$$
where our prior proposal distribution is $Q(\boldsymbol{\tau}'_\mathcal{I}|\boldsymbol{\tau}_{-\mathcal{I}}, \sigma_u^2, \boldsymbol{y}) = P(\boldsymbol{\tau}'_\mathcal{I}|\boldsymbol{\tau}_{-\mathcal{I}}, \sigma_u^2)$. By considering
$$
\begin{aligned}
  P( \boldsymbol{\tau}'_ \mathcal{I}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2, \boldsymbol{y}) &= \frac{P( \boldsymbol{\tau}'_ \mathcal{I}, \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2, \boldsymbol{y})}{P( \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2, \boldsymbol{y})} 
  \\
  &= \frac{P(\boldsymbol{y}| \boldsymbol{\tau}'_ \mathcal{I}, \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2)P(\boldsymbol{\tau}'_ \mathcal{I}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2)P(\boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2)}{P(\boldsymbol{y}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2) P(\boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2)}
  \\
  &= \frac{\overbrace{P(\boldsymbol{y}| \boldsymbol{\tau}'_ \mathcal{I}, \boldsymbol{\tau}_ {-\mathcal{I}})}^{\text{Conditionally independent}} P(\boldsymbol{\tau}'_ \mathcal{I}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2)}{P(\boldsymbol{y}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2)}
  \\
  &= \frac{P(\boldsymbol{y}_\mathcal{I}| \boldsymbol{\tau}'_ \mathcal{I})P(\boldsymbol{y}_\mathcal{-I}| \boldsymbol{\tau}_ \mathcal{-I}) P(\boldsymbol{\tau}'_ \mathcal{I}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2)}{P(\boldsymbol{y}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2)}
\end{aligned}
$$
and equally
$$
  P( \boldsymbol{\tau}_ \mathcal{I}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2, \boldsymbol{y}) 
  = 
  \frac{P(\boldsymbol{y}_\mathcal{I}| \boldsymbol{\tau}_ \mathcal{I})P(\boldsymbol{y}_\mathcal{-I}| \boldsymbol{\tau}_ \mathcal{-I}) P(\boldsymbol{\tau}_ \mathcal{I}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2)}{P(\boldsymbol{y}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2)}
$$
we can rewrite the acceptance probability as
$$
\begin{aligned}
  \alpha(\boldsymbol{\tau}_ \mathcal{I}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2, \boldsymbol{y}) &= \text{min} \left( 1, \frac{{P(\boldsymbol{y}_\mathcal{I}| \boldsymbol{\tau}'_ \mathcal{I})P(\boldsymbol{y}_\mathcal{-I}| \boldsymbol{\tau}_ \mathcal{-I}) P(\boldsymbol{\tau}'_ \mathcal{I}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2)}/{P(\boldsymbol{y}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2)}}
  {{P(\boldsymbol{y}_\mathcal{I}| \boldsymbol{\tau}_ \mathcal{I})P(\boldsymbol{y}_\mathcal{-I}| \boldsymbol{\tau}_ \mathcal{-I}) P(\boldsymbol{\tau}_ \mathcal{I}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2)}/{P(\boldsymbol{y}| \boldsymbol{\tau}_ {-\mathcal{I}}, \sigma_u^2)}}
  \frac{P(\boldsymbol{\tau}_\mathcal{I}|\boldsymbol{\tau}_{-\mathcal{I}}, \sigma_u^2)}{P(\boldsymbol{\tau}'_\mathcal{I}|\boldsymbol{\tau}_{-\mathcal{I}}, \sigma_u^2)} 
  \right)
  \\
  &= \text{min} \left( 1, \frac{P(\boldsymbol{y}_\mathcal{I}| \boldsymbol{\tau}'_ \mathcal{I})}{P(\boldsymbol{y}_\mathcal{I}| \boldsymbol{\tau}_ \mathcal{I})} 
  \right),
\end{aligned}
$$
which is the minimum of 1 and the ratio of likelihoods conditioned on the proposed values and the old values.


## e) Implementation
In this section we implement an MCMC sampler for the posterior $P(\boldsymbol{\pi}, \sigma_u^2| \boldsymbol{y})$. For the conditional prior, $P(\tau_t | \boldsymbol{\tau}_{-t}, \sigma_u)$, we use an MH steps, and for $\sigma_u$ we use Gibbs steps. We assume $\alpha = 2$ and $\beta = 0.05$ for the response given in Equation \@ref(eq:1response), and the initial value for $\sigma_u^2=0.1$. Initial values for $\tau$ are drawn from a standard normal distribution and the number of iterations is $N=50'000$.
```{r 1Functions, options}
link = function(tau){
  # Expit link
  return(exp(tau)/(1+exp(tau)))
}

logbin = function(n, y, tau){
  # Remake and use this
  return(y*log(1+exp(-tau)) - (n-y)*log(1+exp(tau)))
}

acceptRatio = function(n, y, tauProp, tau){
  # Confirmed faster. N=1000: 4.7 vs 3.81
  return(exp(y*(tauProp - tau) + n*log((1+exp(tau))/(1+exp(tauProp)))))
}


mhRW = function(tau, sigma, yt, t, normVec=NA){
  if (t==1){
    mu_ab = tau[2]
    sigma_aa = sigma
  }
  else if (t==366){
    mu_ab = tau[365]
    sigma_aa = sigma
  }
  else{
    mu_ab = 1/2 * (tau[t-1] + tau[t+1])
    sigma_aa = sigma/2
  }
  # prop_tau = rnorm(1,mean=mu_ab, sd=sqrt(sigma_aa))
  # prop_tau = normVec[t]*sigma + mu_ab
  prop_tau = normVec*sigma + mu_ab
  n = ifelse(t==60,10,39)
  ratio = acceptRatio(n, yt, prop_tau, tau[t])
  if (runif(1) < min(c(1,ratio))){
    return(list(tau=prop_tau, accepted=1))
  }
  else{return(list(tau=tau[t], accepted=0))}
}


mcmcIndivid = function(N, dt, sigma0=0.1){
  # Allocate memory
  Ttot = 366
  tau = matrix(NA, nrow=N, ncol = Ttot)
  sigma = numeric(length = N)
  tau_i = numeric(length = Ttot)
  normMat = matrix(rep(rnorm(Ttot), N), nrow = N, ncol=Ttot)
  normVec = normMat[1,]
  
  # Find init vals
  tau[1,] = rnorm(Ttot) # init tau drawn from normal distr.
  # tau[1,] = runif(366, -100, 100) # init tau drawn from uniform distr.
  sigma[1] = sigma0
  
  # Make Q matrix
  Q=matrix(0, nrow = Ttot, ncol = Ttot)
  diag(Q)=2
  Q[c(1, length(Q))]=1
  Q[abs(row(Q) - col(Q)) == 1] <- -1
  
  # Run mcmc for N iterations
  accepted = 0
  for (i in 2:N){
    tau_i = tau[i-1,]
    sigma_i = sigma[i-1]
    for (t in 1:Ttot){
      # rtemp= mhRW(tau_i, sigma_i, dt$n.rain[t], t)
      rtemp= mhRW(tau_i, sqrt(sigma_i), dt$n.rain[t], t, normVec[t])
      tau[i,t] = rtemp$tau
      accepted = accepted + rtemp$accepted
    }
    normVec = normMat[i,]
    # Squared diff. of tau vec.
    tQt = sum((tau[i,-Ttot] - tau[i,-1])^2) # this sim tau vals.
    # tQt = sum((tau[i-1,-366] - tau[i-2,-1])^2) # prev sim tau vals.
    
    # Gibbs step (Draw from IG)
    sigma[i] = 1/rgamma(1, 2 + (Ttot-1)/2, 0.05 + 0.5*tQt) # Gibbs inline

    # if (i%%(N/10)==0){
    #   print(i/N*100)
    #   print(accepted/(i*366))
    # }
  }
  return(list(tau=tau, sigma=sigma, accProb = accepted/(N*Ttot)))
}
```

```{r 1runMCMC, options}
set.seed(321)
N = 50000
ptm = proc.time()
results = mcmcIndivid(N, rain)
time = proc.time() - ptm
```

```{r 1eDisplay, fig.width=7, fig.height=8, out.width="80%", fig.cap="Traceplot, autocorrelation and histogram plots for $\\pi(\\tau_1), \\pi(\\tau_{201})$ and $\\pi(\\tau_{366})$ from top to bottom."}
CImean = function(p, col="cyan3"){
  c(mean=mean(p), quantile(p, probs = c(0.025,0.975)))
}

plotTAH = function(p, hcol = "cyan3", xlab = "", ylab = ""){
  # Plots trace, autocorr and hist of probs
  plot(p, type="l", xlab = "Iterations", ylab="Probability")
  abline(h=mean(p), col="red")
  acf(p, main="")
  hist(p, nclass=40, prob=T, main="", xlab="Probability")
  abline(v = quantile(p, probs = 0.025), col=hcol)
  abline(v = quantile(p, probs = 0.975), col=hcol)
}

p1 = link(results$tau[,1])
p201 = link(results$tau[,201])
p366 = link(results$tau[,366])

par(mfrow=c(3,3))
plotTAH(p1)
plotTAH(p201)
plotTAH(p366)
```
The vertical red line in the trace plots to the left in Figure \@ref(fig:1eDisplay) show the mean of $\pi(\tau_t)$ over all $50'000$ iterations, for $t = 1, 201, 366$. The horizontal blue lines in the histogram plot show the $95\%$ credible intervals for $\pi(\tau_t)$. The traceplot bares resemblance to that of a random walk for all $\tau_t$. We notice an exponential decrease in autocorrelation and the histogram show that $\tau_{201}$ seem normally distributed and $\tau_1$ and $\tau_{366}$ has a slight right skew.
```{r 1eSigmaRes, fig.width=5, fig.height=8, out.width="80%"}
par(mfrow=c(3,1))
yl = expression(sigma_u^2)
plot(results$sigma, type="l", ylab=yl)
acf(results$sigma)
hist(results$sigma, nclass=100, prob=T)
```

```{r makeTitle, options}
ciAndMean = rbind(CImean(p1[1:N]),
                  CImean(p201[1:N]),
                  CImean(p366[1:N]))
rownames(ciAndMean) = c(1, 201, 366)
ciAndMean
```

```{r 1eEvolution, fig.width=5, fig.height=8, out.width="80%"}
par(mfrow=c(3,1))
plot(link(results$tau[1,]), type = "l")
plot(link(results$tau[N/2,]), type = "l")
plot(link(results$tau[N,]), type = "l")
```